---
title: "Introduction to excelerate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction
This vignette demonstrates how to use the `excelerate` package to create supplementary tables from example datasets.

```{r setup}
library(excelerate)

# Set up temporary directory to save the excel spreadsheets in
temp_dir <- tempdir()
```

## Viewing Example Datasets
For our examples we will use the built-in datasets: iris, ChickWeight, ToothGrowth, and CO2.

```{r}
# Load example datasets
datasets <- list(iris, ChickWeight, ToothGrowth, CO2)
lapply(datasets, head)
```

## Creating Metadata

We will use `append_meta()` to add comments on each of the columns in these dataframes with a description.
This metadata will be presented in the README sheet of the Excel file.

The `append_meta()` function requires the `results` argument to be a dataframe and the `colname_descriptions` argument to be a named character vector where the names are the column names and the values are the descriptions. If the column is a factor then include descriptions for the meaning of each factor level.

```{r}
iris_commented <- append_meta(
  results = iris,
  colname_descriptions = c(
    "Sepal.Length" = "Length of the sepal in cm",
    "Sepal.Width" = "Width of the sepal in cm",
    "Petal.Length" = "Length of the petal in cm",
    "Petal.Width" = "Width of the petal in cm",
    "Species" = "Species of iris"
  )
)

chickweight_commented <- append_meta(
  results = ChickWeight,
  colname_descriptions = c(
    "weight" = "Weight of the chick in grams",
    "Time" = "Time in days",
    "Chick" = "Chick identifier",
    "Diet" = "Diet type"
  )
)

toothgrowth_commented <- append_meta(
  results = ToothGrowth,
  colname_descriptions = c(
    "len" = "Tooth length in mm",
    "supp" = "Supplement type (VC or OJ)",
    "dose" = "Dose of the supplement in mg"
  )
)
co2_commented <- append_meta(
  results = CO2,
  colname_descriptions = c(
    "Plant" = "Identifier for the plant",
    "Type" = "Type of plant (Quebec or Mississippi)",
    "Treatment" = "Treatment type (nonchilled or chilled)",
    "conc" = "Concentration of CO2 in uL/L",
    "uptake" = "CO2 uptake in mg"
  )
)
```

## Generating Supplementary Tables

Next, we demonstrate how to combine sheets into a single Excel file using the `sheet()`, `spreadsheet()` and `excelerate()` functions. In the below example we create two supplementary tables, each containing multiple sheets with metadata and data from the example datasets. These are saved as Excel files in the temporary directory with the names `supplementary_table_1.xlsx` and `supplementary_table_2.xlsx`.

```{r}
supplementary_table_1 <- spreadsheet(
  title = "Results for iris and ChickWeight. 'title' from spreadsheet().",
  path = file.path(temp_dir),
  file = "",
  sheet(
    iris_commented,
    "iris sheet_name",
    "sheet_legend for the iris dataset."
  ),
  sheet(
    chickweight_commented,
    "ChickWeight sheet_name",
    "sheet_legend for the ChickWeight dataset."
  )
)

supplementary_table_2 <- spreadsheet(
  title = "Results for toothgrowth and C02 R supplied datasets.",
  path = file.path(temp_dir),
  file = "",
  sheet(
    toothgrowth_commented,
    "ToothGrowth sheet_name",
    "sheet_legend for the ToothGrowth dataset."
  ),
  sheet(
    co2_commented,
    "CO2 sheet_name",
    "sheet_legend for the CO2 dataset."
  )
)

# Save supplementary tables as .xlsx to temp_dir
excelerate(
  supplementary_table_1,
  supplementary_table_2
)
```

## Inspect the saved Excel files

The Excel files have been saved in a temporary directory that we definied at the
top of this script. You can inspect these files using the following steps:

1. List the files in the temporary directory to make sure they have been 
  created.
2. Open the Excel files directly from R using `browseURL()`.

```{r inspect-files}
# List all files in the temporary directory
list.files(temp_dir, pattern = "\\.xlsx$", full.names = TRUE)

# Open the files to inspect them
# The file name has now appended "S1" and "S2" by default in excelerate()
browseURL(file.path(temp_dir, "S1.xlsx"))
browseURL(file.path(temp_dir, "S2.xlsx"))
```

## Customise the prefix to sheet name, spreadsheet title and filenames

`sheet_template` must contain `{n}` (referring to the table number) and `{l}` (referring to the sheet letter). `spreadsheet_template` must contain `{n}` (referring to the table number).

```{r}
supplementary_table_1 <- spreadsheet(
  title = "Results for iris and ChickWeight. 'title' from spreadsheet().",
  path = file.path(temp_dir),
  file = "iris ChickWeight",
  sheet(
    iris_commented,
    "iris sheet_name",
    "sheet_legend for the iris dataset."
  ),
  sheet(
    chickweight_commented,
    "ChickWeight sheet_name",
    "sheet_legend for the ChickWeight dataset."
  )
)

supplementary_table_2 <- spreadsheet(
  title = "Results for toothgrowth and C02 R supplied datasets.",
  path = file.path(temp_dir),
  file = "",
  sheet(
    toothgrowth_commented,
    "ToothGrowth sheet_name",
    "sheet_legend for the ToothGrowth dataset."
  ),
  sheet(
    co2_commented,
    "CO2 sheet_name",
    "sheet_legend for the CO2 dataset."
  )
)

# Append "1_A_", "1_B_" to each sheet name
# and "Supplementary Table 1" to the legend title and file name
# by default any spaces in the file name will be converted to underscores
excelerate(
  supplementary_table_1,
  supplementary_table_2,
  sheet_template = "{n}_{l}_",
  spreadsheet_template = "Supplementary Table {n}"
)

list.files(temp_dir, pattern = "\\.xlsx$", full.names = TRUE)
browseURL(file.path(temp_dir, "Supplementary_Table_1_iris_ChickWeight.xlsx"))
browseURL(file.path(temp_dir, "Supplementary_Table_2.xlsx"))
```

## Conclusion

This vignette has demonstrated how to create supplementary tables using the 
`excelerate` package in R. We viewed example datasets, attached metadata to 
these datasets, and generated Excel spreadsheets with multiple sheets. The 
`excelerate` package simplifies the process of creating Supplementary Tables 
which can often be a time consuming, manual task in manuscript preparation.
